Monoidal Toposes of Grammars and Semantic Actions

### 2022-08-22 notes.tex
getting a little too speculative

\newcommand{\leftmultimap}{\mathop{\rotatebox[origin=c]{180}{$\multimap$}}}
\newcommand{\Set}{\textrm{Set}}
\newcommand{\Lang}{\textrm{Lang}}
\newcommand{\Grammar}{\textrm{Grammar}}
\newcommand{\Action}{\textrm{Action}}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}

@@

\begin{definition}
  A notion of tree with binding $\mathcal T$ is a cartesian multicategory all of
  whose objects are exponentials $o \to^n o$ for some distinguished
  object $o$.
\end{definition}

\begin{definition}
  Given a notion of tree with binding $\mathcal T$, the topos of
  grammars is simply $\Set^{|Arr(\mathcal T)|}$.

  But there are many sorts here...
\end{definition}

Again the interesting question is what the appropriate notion of
``separating conjunction'' is for this notion of grammar. Following
the pattern thus far, the categorical structure we want on the
category of grammars should be given by a pseudo-algebra of the notion
of theory the grammars are defined over. An algebra of a 2nd-order
theory is given by an exponentiable object in a category with
cartesian products with an interpretation of all the operations.

In this case this means that we want to interpret the 2nd-order
operations as functors whose domain is \emph{a category of
endofunctors}.

An example should be illuminating. One of the canonical examples of a
second-order theory is the untyped lambda calculus. This is a single
sorted theory with two operations :

\begin{enumerate}
\item \texttt{app : (0, 0)}
\item \texttt{lam : (1)}
\end{enumerate}

That is, application is applied to two subterms with no introduced
bindings, whereas lambda is applied to a subterm with an additional
binding introduced.  As morphisms in a 2nd order algebraic theory
viewed as a cartesian multicategory these are in the hom sets:

\begin{enumerate}
\item \texttt{app : Hom(o, o; o)}
\item \texttt{lam : Hom(o -> o; o)}
\end{enumerate}

To give the topos of grammars $G(\mathcal T)$ an algebra structure, we
define a functor of two arguments for \texttt{app}:
\[ \textrm{app}(g_1,g_2)_t = \sum_{\texttt{app}(t_1,t_2)} g_1 t_1 \times g_2 t_2 \]
and a functor whose argument is an endofunctor for \texttt{lam}:
\[ \textrm{lam}(F)_t = \sum_{\texttt{lam}(x.s)} F(Y x)_s \]
this doesn't quite make sense...
%% Something something second-order algebraic theory..

Intuitively, the functor takes in a grammar for the variable and
outputs a grammar over terms with that free variable in them.

>>>>

\newcommand{\leftmultimap}{\mathop{\rotatebox[origin=c]{180}{$\multimap$}}}
\newcommand{\Set}{\textrm{Set}}
\newcommand{\FinSet}{\mathbb F}
\newcommand{\Cat}{\textrm{Cat}}
\newcommand{\Lang}{\textrm{Lang}}
\newcommand{\Grammar}{\textrm{Grammar}}
\newcommand{\Action}{\textrm{Action}}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}

@@

\begin{definition}
  A notion of tree with binding $\mathcal T$ is a cartesian
  multicategory all of whose objects are exponentials $o \to^n o$ for
  some distinguished generator object $o$.
\end{definition}

The ``trees'' themselves are the morphisms of the multicategory of the
form
\[ \mathcal T(\cdot; o) \]

The ``constructors'' of this notion of are given by the morphisms of
the multicategory, in which can all be given in the form:
\[ c \in \mathcal T((o \to^{n_1} o), \ldots_{p}; o) \]
A constructor $c$ of this form has $p$ subterms each of which has
bindings for $n_i$ new variables.

So what category do our grammars live in?
%
In analogy with the string and tree cases, our category of grammars
should be some kind of \emph{pseudo}-algebra of the theory in $\Cat$.
%
However, unlike the string and tree case, even ``single sorted''
second-order theories behave more like multi-sorted theories, in that
a model of a second-order theory needs to interpret not just the
generator $o$, but also all $n-ary$ exponentials $o \to^n o$ as
categories.
%
This suggests we define for each natural number $n$, a category of
grammars over trees with $n$ variables:
\[ \Grammar(n) = \Set^{\mathcal T(o,\ldots_n;o)} \]
This varies functorially in $n$, defining a functor
\[ \Grammar(-) : \FinSet \to \Cat \]
where $\FinSet$ is a skeleton of the category of finite sets and
arbitrary functions, i.e., the action of ``renaming'' the variables.

Then we get for each operation $f \in \mathcal T((o^{n_1} \to o,
\ldots_p; o^m \to o))$ a functor $\Grammar(f) : \Grammar(n_1) \times
\cdots_p \to \Grammar(m)$ defined as a kind of separating conjunction:
\[ \Grammar(f)(g_1,\ldots_p)(t) = \sum_{f(\vec x_1. s_1,\ldots_p) = t} (g_1(\vec x_1.s_1)) \times \cdots \]
Does that even make sense??

Maybe we can combine these into one using Grothendieck construction?
Not sure it has a nice internal language though.
\[ \Grammar = \int_{[n] : \FinSet} \Grammar(n) \]

%% Then we can construct a category of grammars for trees with binding as
%% the Grothendieck construction:
%% Where objects are pairs of a finite cardinal and a grammar over trees
%% with that many variables, with a morphism being a pair of a renaming
%% and then a parse transformer between a grammar and a ``renamed''
%% grammar.

%% Again the interesting question is what the appropriate notion of
%% ``separating conjunction'' is for this notion of grammar. Following
%% the pattern thus far, the categorical structure we want on the
%% category of grammars should be given by a pseudo-algebra of the notion
%% of theory the grammars are defined over. An algebra of a 2nd-order
%% theory is given by an exponentiable object in a category with
%% cartesian products with an interpretation of all the operations.

%% In this case this means that we want to interpret the 2nd-order
%% operations as functors whose domain is \emph{a category of
%% endofunctors}.

%% An example should be illuminating. One of the canonical examples of a
%% second-order theory is the untyped lambda calculus. This is a single
%% sorted theory with two operations :

%% \begin{enumerate}
%% \item \texttt{app : (0, 0)}
%% \item \texttt{lam : (1)}
%% \end{enumerate}

%% That is, application is applied to two subterms with no introduced
%% bindings, whereas lambda is applied to a subterm with an additional
%% binding introduced.  As morphisms in a 2nd order algebraic theory
%% viewed as a cartesian multicategory these are in the hom sets:

%% \begin{enumerate}
%% \item \texttt{app : Hom(o, o; o)}
%% \item \texttt{lam : Hom(o -> o; o)}
%% \end{enumerate}

%% To give the topos of grammars $G(\mathcal T)$ an algebra structure, we
%% define a functor of two arguments for \texttt{app}:
%% \[ \textrm{app}(g_1,g_2)_t = \sum_{\texttt{app}(t_1,t_2)} g_1 t_1 \times g_2 t_2 \]
%% and a functor whose argument is an endofunctor for \texttt{lam}:
%% \[ \textrm{lam}(F)_t = \sum_{\texttt{lam}(x.s)} F(Y x)_s \]
%% this doesn't quite make sense...
%% %% Something something second-order algebraic theory..

%% % Examples: Schemes' let vs letrec vs let*

%% Intuitively, the functor takes in a grammar for the variable and
%% outputs a grammar over terms with that free variable in them.
